---
title: "Linear Regression Lab"
format: html
editor: visual
---

```{r}
#| message: false
#| warning: false

# Import all libraries
library(knitr)
library(GGally)
library(janitor)
library(tidyverse)
library(ggthemes)

# Use a nicer theme
theme_set(theme_few())
```

# Setup: the data

The data for this lab come from a lexical decision experiment (Baayen et al. 2006). Twenty-one participants saw 79 words -- the names of animals and plants -- and decided as quickly as possible whether each was a real word. We recorded their reaction times (log-transformed, in the `RT` column).

The reaction times are stored in `lexdec_wide.csv` in *wide* format (one column per word). Word-level properties -- frequency, morphological family size, length, and class -- live in `wordinfo.csv`. The chunk below reshapes the reaction times into *long* format and joins in the word properties, giving us a single tidy data frame. (You don't need to change anything here -- just run it.)

```{r}
#| message: false

base_url <- paste0("https://raw.githubusercontent.com/SocialInteractionLab/",
                   "LINGUIST245B/refs/heads/master/assets/data/")

df.lexdec_wide <- read_csv(paste0(base_url, "lexdec_wide.csv"))
df.wordinfo    <- read_csv(paste0(base_url, "wordinfo.csv"))

df.lexdec <- df.lexdec_wide %>%
  pivot_longer(cols = -c(Subject, Sex, NativeLanguage),
               names_to = "Word", values_to = "RT") %>%
  left_join(df.wordinfo, by = "Word") %>%
  clean_names() %>%
  rename(word_length = length)

head(df.lexdec)
```

# I. Simple linear regression and prediction

## 1. Explore the data

The plot below shows all pairwise relationships among the numeric variables. (You don't need to change this chunk.)

```{r}
#| message: false

df.lexdec %>%
  select(rt, frequency, family_size, word_length) %>%
  ggpairs(progress = FALSE)
```

Looking at the plot, briefly describe one relationship that stands out to you.

YOUR ANSWER HERE: Reaction time seems to decrease as frequency increases

## 2. Fit and interpret a simple regression

You want to test the relationship between a word's frequency (`frequency`) and how quickly people recognize it (`rt`).

**a)** Make a scatterplot of `rt` (y-axis) against `frequency` (x-axis) with point transparency 0.2, and save it to an object called `plot.rt_frequency`. The last line of the chunk displays it.

```{r}

### CODE HERE
plot.rt_frequency <- ggplot(df.lexdec, aes(x = frequency, y = rt)) +
  geom_point(alpha = 0.2)

plot.rt_frequency

```

**b)** Fit a linear regression predicting `rt` from `frequency`, and look at the model summary.

```{r}
m1 <- lm(rt ~ 1, data = df.lexdec)

m2 <- lm(rt ~ frequency, data = df.lexdec)

summary(m2)
anova(m1, m2)

```

**c)** Is frequency a significant predictor of reaction time? What is the r-squared value, and what does it tell you about the model?

YOUR ANSWER HERE: Yes. Frequency is a significant predictor of reaction time (p \< .001; more exactly about 1.03e-20). The r-squared value is about 0.0528, which means frequency explains about 5.3% of the variation in reaction time. So there is a real relationship, but frequency by itself only explains a small part of why reaction times differ.

## 3. Prediction and model fit

**a)** Use your model to predict `rt` for every row, storing the result in a new `predictions` column of `df.lexdec`.

> Hint: take a look at the `predict()` function.

```{r}

### CODE HERE
df.lexdec <- df.lexdec %>%
  mutate(predictions = predict(model.rt_frequency, newdata = df.lexdec))
```

The chunk below overlays those predictions (red) on the observed data, with black segments showing the residuals -- the gap between each observed and predicted value. (You don't need to change this chunk.)

```{r}
plot.rt_frequency +
  geom_segment(data = df.lexdec,
               aes(xend = frequency, yend = predictions), color = "black") +
  geom_point(data = df.lexdec,
             aes(y = predictions), color = "red")
```

Squaring and summing those residuals gives the *residual* sum of squares; comparing it to the *total* sum of squares gives r-squared. The function below computes it (you don't need to change it):

```{r}
r_squared <- function(Y_true, Y_pred) {
  ss_residual <- sum((Y_true - Y_pred)^2)
  ss_total    <- sum((Y_true - mean(Y_true))^2)
  1 - ss_residual / ss_total
}
```

**b)** Use `r_squared()` to compute r-squared for your predictions, and confirm it matches the value reported by `summary()` in 2b.

```{r}

### CODE HERE
r_squared <- function(Y_true, Y_pred) {
  ss_residual <- sum((Y_true - Y_pred)^2)
  ss_total    <- sum((Y_true - mean(Y_true))^2)
  1 - ss_residual / ss_total
}

r_squared(df.lexdec$rt, df.lexdec$predictions)
summary(model.rt_frequency)$r.squared


```

**c)** In theory, what are the minimum and maximum values r-squared can take on **training data** with a linear regression, and why? What does an r-squared of 0 mean?

> YOUR ANSWER HERE: On training data, r-squared can range from 0 to 1. It cannot be below 0 here because the regression model includes an intercept, so it does at least as well as predicting the mean. An r-squared of 0 means the model explains none of the variance in rt beyond the mean.

# II. Controlling for a confound

In psycholinguistics, as in much of psychology, people often run regressions to assess the relationship between two variables while "controlling" for others. But which variables we control for is a substantive decision. Here we'll see what controlling for a variable can do.

A word's **morphological family size** (`family_size`) is the number of morphologically related words -- derived forms and compounds -- it belongs to (log-transformed).

## 4. Family size, with and without frequency

**a)** Fit a simple linear regression predicting `rt` from `family_size` and look at the summary. Is family size a significant predictor of reaction time? Briefly, what do the intercept and slope mean?

```{r}

### CODE HERE

lm(rt ~ family_size, df.lexdec) |>
summary()
```

> YOUR ANSWER HERE:Yes, if the p-value for family_size is below .05. The intercept is the predicted rt when family_size = 0. The slope is the change in predicted rt for a 1-unit increase in family_size.

Morphological family size is strongly correlated with word frequency -- words with big families also tend to be high-frequency -- and frequency is itself a powerful predictor of recognition speed.

**b)** Now fit a regression predicting `rt` from BOTH `family_size` and `frequency`. Is family size still a significant predictor? What could explain the change, if there is one?

```{r}

### CODE HERE
lm(rt ~ family_size + frequency, df.lexdec) |>
summary()
```

> YOUR ANSWER HERE: Family size is not still significant after adding frequency (p ≈ .095). A likely reason is that family size and frequency overlap, so once frequency is included, family size explains little additional variance.

# III. Interactions

So far we've treated the frequency effect as if it were the same for every reader. But the speed-up that frequency buys might depend on who is reading. In this section we'll ask whether the effect of `frequency` on `rt` is *moderated* by a participant's language background (`native_language`: `English` vs. `Other`).

## 5. Visualize the interaction

Make a scatterplot of `rt` against `frequency`, add a linear regression line, and facet by `native_language`.

> Hint: `geom_smooth(method = "lm")` and `facet_wrap()`.

```{r}

### CODE HERE

ggplot(df.lexdec, aes(x = frequency,
                      y = rt,
                      color = native_language)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ native_language)


## ggplot(df.lexdec, aes(x = frequency,
##                    y = rt,
##                    color = native_language)) +
### facet_wrap(~ native_language)
## geom_smooth(method ='lm')
```

Another way to see the same thing is to chop `frequency` into thirds (terciles) and compare language groups within each. The chunk below builds the terciles and plots them (you don't need to change it):

```{r}
cut_terciles <- function(x) {
  cut(x,
      breaks = quantile(x, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
      include.lowest = TRUE)
}

df.lexdec <- df.lexdec %>%
  mutate(frequency_terciles = cut_terciles(frequency))

ggplot(df.lexdec, aes(x = native_language, y = rt, color = frequency_terciles)) +
  geom_boxplot() +
  facet_wrap(~ frequency_terciles)
```

## 6. Test the interaction

Fit two linear models with `rt` as the outcome: one with `frequency` and `native_language` but **no** interaction, and one that **adds** their interaction. Compare the two models with `anova()`.

```{r}

### CODE HERE
m1 <- lm(rt ~ native_language + frequency, df.lexdec)
m2 <- lm(rt ~ native_language + frequency, df.lexdec)
anova(m1,m2)
```

Is the interaction term doing real work? How does that match what you saw in the plots in Problem 5?

> YOUR ANSWER HERE: The interaction improves the model, which matches the plots because the frequency slopes differ by language background.

## 7. Interpret the interaction

`native_language` is already binary, so we can make an explicit 0/1 dummy variable: 0 for native English speakers, 1 otherwise. The chunk below builds it and fits the interaction model (you don't need to change it):

```{r}
df.lexdec2 <- df.lexdec %>%
  mutate(nonnative = as.integer(native_language == "Other"))

interaction_model <- lm(rt ~ 1 + frequency + nonnative + nonnative:frequency,
                        data = df.lexdec2)
summary(interaction_model)
```

The model is:

$$rt_i = \beta_0 + \beta_1 \cdot frequency_i + \beta_2 \cdot nonnative_i + \beta_3 \cdot (nonnative_i \cdot frequency_i) + \epsilon_i$$

**a)** What does each of $\beta_0$, $\beta_1$, $\beta_2$, and $\beta_3$ mean substantively -- in terms of frequency, reaction time, and language background?

YOUR ANSWER HERE: β0 is the predicted rt for native English speakers when frequency = 0. β1 is the effect of frequency on rt for native English speakers. β2 is the difference in predicted rt between non-native and native speakers when frequency = 0. β3 is how much the frequency effect changes for non-native speakers compared with native speakers.

**b)** Based on the summary, is there evidence of moderation -- does the frequency effect differ by language background? Which coefficient tells you, and which direction does it go?

> YOUR ANSWER HERE: Yes. The interaction coefficient β3 tells us this. It is negative, so the frequency effect is more negative for non-native speakers.

# A note on independence

Throughout this lab we have fit ordinary linear models with `lm()`, which assume that every row of the data is an independent observation. That assumption is **not met** here: each of our 21 participants contributed a response to all 79 words, so the rows are clustered both by participant and by word. Next week we will see how *mixed-effects models* (`lmer`) let us model that structure directly, instead of pretending it isn't there.

# Session info

```{r}
sessionInfo()
```
